---
title: "Class 08: Breast Cancer Analysis Mini Project"
author: "Ashley Mazon (PID:A17478903)"
format: pdf
toc: TRUE
---

## Background 

The goal of this mini-project is for you to explore a complete analysis using the unsupervised learning techniques covered in class. You’ll extend what you’ve learned by combining PCA as a preprocessing step to clustering using data that consist of measurements of cell nuclei of human breast masses. This expands on our RNA-Seq analysis from last day.

The data itself comes from the Wisconsin Breast Cancer Diagnostic Data Set first reported by K. P. Benne and O. L. Mangasarian: “Robust Linear Programming Discrimination of Two Linearly Inseparable Sets”.

Values in this data set describe characteristics of the cell nuclei present in digitized images of a fine needle aspiration (FNA) of a breast mass.

## Data Import 

Input data file is saved into my Project directory as a CSV

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
head(wisc.df)

```


The first column `diagnosis` is the expert opinion on the sample (i.e. patient FNA)
```{r}
wisc.df$diagnosis
```


-1 is used to remove the diagnosis column, which is not needed right now, but needed later as a vector

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df[,1]
```

Lastly, explore and get familiar 

>Q1. Q1. How many observations are in this dataset?

There are `r nrow(wisc.data)` observations/patients in the dataset.


>Q2. How many of the observations have a malignant diagnosis?

There are 212 observations that have a malignant diagnosis.

```{r}
table(wisc.df$diagnosis)
```

>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
colnames(wisc.data)
length(grep("_mean", colnames(wisc.data)))
```
## Principal Coordinate Analysis 

The `prcomp()` function to do PCA has a `scale=FALSE`default. In general we nearly always want to set this to TRUE so our analysis is not dominated by columns/variables in our data set that have high standard deviation and mean when compared to others just because the units of measurement are on different scales 

```{r}
wisc.pr <- prcomp( wisc.data,scale=TRUE)
summary(wisc.pr)
```

The main PC result figure is called a "score plot" or a "PC polt" or "ordination plot"...

```{r}
library(ggplot2)

ggplot(wisc.pr$x)+
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```



> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

44.27% of the original variance is captured by PC1

```{r}
summary(wisc.pr)
```


> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 principal components are required to describe at least 70% of the data 

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

About 8 principal components are requied to describe at least 90% of the original variance in the data 


## Interpreting PCA Results

Create a biplot of the `wisc.pr` using the `biplot()` function 

```{r}
biplot(wisc.pr)
```

> Q7. What stand out to you about this plot? Is it easy or difficult to understand and why?

There are too many plots on this plot, there is no scaling of values to remove this noise. 

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots? 

These plots show a distinct difference in clustering between the two colors. 

```{r}
col <- ifelse(diagnosis == "M", "red", "blue")
plot(wisc.pr$x[,c(1,3)], col = col , 
     xlab = "PC1", ylab = "PC3")
```


## PCA Scree-plot 


A plot of how much variance each PC captures. We can get this from `wisc.pr$sdev` or from the output of `summary(wisc.pr)`

```{r}
var.tbl <- summary(wisc.pr)
head(var.tbl$importance)

```

```{r}
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

Variance explained by each principal component: pve
```{r}
pve <- pr.var / sum(pr.var)

```

```{r}
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

An alternative scree plot can be made 

```{r}
screeplot <- barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

> Q9. For the first principal component, what is the component of the loading vector (i.e. `wisc.pr$rotation[,1]`) for the feature `concave.points_mean`?

```{r}
wisc.pr$rotation["concave.points_mean", 1]

```

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

5 principal components are required to explain 80% of the variance of the data 




## Hierarchical clustering 

Just clusting the original data is not very informative or helpful 

```{r}
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust (data.dist)
```

```{r}
plot (wisc.hclust)
```
```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(cutree(wisc.hclust, k=4))
```
```{r}
table(wisc.hclust.clusters,diagnosis)
```




## Combining methods (PCA and Clustering)

Clustering the original data was not very productive. The PCA results looked promising. Here we combined these methods by clustering from our PCA results. In other words "clustering in PC space"

```{r}
dist.pc <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(dist.pc, method = "ward.D2")
```


> Q11. Using the `plot()` and `abline()` functions, what is the height at which the clustering model has four clusters?

height=35 is when the clustering model of `wisc.pr.hclust` has four clusters. 

View the tree... 

```{r}
plot(wisc.pr.hclust)
abline(h=35, col="red")
```

To get out clustering membership vector (i.e. out main clustering result) we "cut" the tree at a desired height or to yield a desired number of "k" 

```{r}
grps <- cutree(wisc.pr.hclust, h=70)
table(grps)
```
How does this clustering grps compare to the expert diagnosis 
```{r}
table(grps, diagnosis)
```
> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10? 

Yes, because there is less split of the data. When there is two clusters there is a split in data where there is a majority of M in one and B in the other. 

> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

Using `ward.D2` provides a clear separation between M and B samples that were both balanced. Overall, M and B were accurately grouped and a strong alignment in each cluster can be observed. 

## Combining Methods 

Create a graph of PC1 and PC2 
```{r}
col <- ifelse(diagnosis == "M", "red", "black")
plot(wisc.pr$x[,1:2], col=col)

```

Re-order the factor and re plot 

```{r}
g <- as.factor(grps)
levels(g)
```
```{r}
g <- relevel(g,2)
levels(g)
```
```{r}
plot(wisc.pr$x[,1:2], col=g)
```

Cut this model into two clusters and assign to `wisc.pr.hclust.clusters`

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
table(wisc.pr.hclust.clusters, diagnosis)
```
> Q15. How well does the newly created model with four clusters separate out the two diagnoses?

There is a distinct and balaned separation of the four clusters. Cluster 1 consist of majority M samples and cluster 2 consists of majority B samples. 

> Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

The k means model separates the diagnoses well however there is some overlap. The heiarchical model shows a distinct separation but with four clusters instead of 2, thus fragmenting the separation. 


Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity? 

Hierarchical clustering is both higher in sensitivity and specificity than the k means model. This was concluded as there were greater benign and malignant samples that were correctly grouped. 

Sensitivity: TP/(TP+FN)
Specificity: TN/(TN+FN)

## 7. Prediction 

We can use out PCA model for prediction with new input patient samples. 

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize patient 2 due to closer clusering associated with a malignant disease diagnosis. They are a higher risk group for malignant disease. 

